---
title: "lab-exercise"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Outline

1. Get Data and Libraries
2. Introducing the dataset
3. Framing a research  question
4. Preliminary data exploration

<br>

## Get Data and Libraries

We will use data from the [`palmerpenguins`](https://github.com/allisonhorst/palmerpenguins) package for the first few sections.

```{r get data}
# install.packages("remotes")
remotes::install_github("allisonhorst/palmerpenguins")
penguins <- palmerpenguins::penguins
```


<br>

## Introducing the dataset
The palmer penguins data set contains data on the features of three different penguin species. You can have a look at the different variables using the function `head()`:

```{r}
head(penguins)
```

Before even starting to explore the data set you should be aware of the different data types you are working with. In the penguins data set we can find two different types of data: discrete or categorical data and continuous data. Discrete data are classified in penguins as factors (e.g. species, island and sex). The other variables are continuous and belong either to the class double or integer. 

`r shiny::icon("star")` Tip 1: Check for missing values (NAs) or potential outliers using the `summary()` function. 
```{r}
summary(penguins)
```


## Framing a research question

We want to use the palmer penguins data to show how to answer some common questions, applied in a SES context. 
The basic research question driving the exploration is: What are differences / similarities between the states (penguin species)? Are there spatio-temporal patterns / changes? 
Before starting the analysis it is always useful to do some basic data exploration / visualization exercises to get a "feeling" for your data -> How does it look like? 

For this we install the tidyverse packages, which includes various other packages useful for visualization (ggplot2), data wrangling (dplyr, tidyr), string manipulation (stringr) and others. 

```{r, warning = FALSE, message = FALSE}
#install.packages(tidyverse)
library(tidyverse)

ggplot(data = penguins, aes(x = penguins$species, y = penguins$bill_length_mm))+
  geom_boxplot()+
  xlab("Species")+ ylab("Bill length (mm)")

ggplot(data = penguins)+
  geom_histogram(aes(x = bill_length_mm,  fill = species))+
  xlab("Bill length (mm)") + ylab("Frequency") +
  scale_fill_manual(values = c("black", "orange", "skyblue"))+
  labs(fill = "Species")
  
```

<mark>**Exercise 1: Now you can explore the other variables on your own. Try to figure out which penguin species has the highest body mass (in g). **</mark>

## Preliminary data exploration

Following the steps outlined in Chapter 18 (`SES Problems and Questions`):

- Identifying key features (mean, variance, normality/distribution of the data)

Now that we have explored and visualized the data set a bit, we want to identify key features of the different species, to summarise the data set. R provides several functions for this task. Here we calculate the mean bill length per species and its standard deviation. 

`r shiny::icon("star")` Tip 2: If you calculate a mean you should always provide a measure for the variation within the data, e.g. the standard deviation or variance. Without the variation a mean is meaningless. 


```{r}
mean(penguins$bill_length_mm, na.rm = TRUE)
sd(penguins$bill_length_mm, na.rm = TRUE)
```

Now we know the mean bill length and its sd across all three penguin species, will those values change if we calculate the same measures per species? 
We are using the tidyverse syntax in the following, you could do the same calculations using base R, it´s just a matter of convenience...

```{r}
penguins%>%
  group_by(species)%>%
  summarise("mean_bill_length" = mean(bill_length_mm, na.rm = TRUE),
            "sd_bill_length" = sd(bill_length_mm, na.rm = TRUE))
```

For non-parametric and discrete data it is more common to use parameters such as the median or quartiles. It is also common to use the median in case of skewed data

<mark>**Exercise 2: Calculate the median bill length per species and compare it the mean values. Is there a difference? Can you explain why there is a difference (or not)? **</mark>

Another part of descriptive statistics is to identify correlations between different variables. 

`r shiny::icon("star")` Tip 3: Correlation ≠ Causation!

We first visually explore the relationship between two variables: 
```{r, warning = FALSE}
ggplot(data = penguins, aes(x = bill_length_mm, y = bill_depth_mm))+
  geom_point()+
  labs(x = "Bill length (mm)", y = "Bill depth (mm)")
```

The points scatter over the entire plot, but it seems like there is a slightly positive correlation between bill length and bill depth. We can validate this by using a correlation test as Pearson´s correlation coefficient test, this test is a parametric test, so we have to check that both our variables are normally distributed (their histograms should have a bell-shape). We already did this for the bill length, so let´s check the bill depth 

`r shiny::icon("star")` Tip 4: Always check the assumptions of your test, before applying any statistics! Common assumptions are normal distribution of the data or variance homogeneity. 

```{r, message = FALSE}
ggplot(data = penguins)+
  geom_histogram(aes(x = bill_depth_mm, fill = species))+
  xlab("Bill depth (mm)") + ylab("Frequency")+
  scale_fill_manual(values = c("black", "orange", "skyblue"))+
  labs(fill = "Species")
```

The individual species have a bell-shaped distribution regarding the bill length. For the bill depth only two species have a bell-shaped distribution, the Chinstraps have a rather uniform distribution. 
For the normally distributed species we now apply Person´s r to test for a correlation between the two variables: 
```{r}
adelie <- penguins%>%
  filter(species == "Adelie")
cor.test(x = adelie$bill_length_mm, y = adelie$bill_depth_mm)

```
So this output tells us that R is confident that there is a correlation (p-value), but it is a rather weak correlation (the correlation coefficient is closer to zero than to 1 or -1). 


<mark>**Exercise 3: Calculate the correlation between bill length and bill depth for the other two species. Remember to check for normality of both variables, if one of them is not normally distributed you have to choose another test (just google for non-parametric correlation test). **</mark>

- Comparison between different categories

If you want to investigate the difference between two (or more) groups you usually compare the mean of the groups. One famous test for this purpose is the t-Test. It has the Null-hypothesis that the true difference of the between two groups is zero. So if the p-value of the output is smaller than 0.05, we can reject the Null-Hypothesis and conclude that there is a significant difference in the mean of both groups. 
There are several other test for group comparisons aimed at different types of data, e.g. the t-Test is suitable for normally distributed data, if you have non-parametric data you have to use another test (see page 260 in the book). 
If you want to compare more than two groups you have to use other tests again. For normal distributed data you can use the ANOVA, for non-parametric data the Kruskal-Wallis or the Chi-squared test can be used. 

`r shiny::icon("star")` Tip 5: Decision trees can help you find the correct test / analysis for your data and research question. For example this one: [`decision tree`](https://www.central7.net/wp-content/uploads/2015/09/stats_flow_chart_v2014.gif). 

In this lab we want to identify if there are differences between the different penguin species regarding their bill length. This means we compare more than two groups, to use the ANOVA we first have to check if all data is normally distributed and if the variance are homogeneous. 

Instead of checking the assumption of normal distribution visually we use this time the Shapiro - Wilks Test: 

```{r}
shapiro.test(penguins[penguins$species == "Adelie", ]$bill_length_mm)
shapiro.test(penguins[penguins$species == "Chinstrap", ]$bill_length_mm)
shapiro.test(penguins[penguins$species == "Gentoo", ]$bill_length_mm)

```
The bill length is normally distributed for the Adelie and Chinstrap penguins, but not for the Gentoos. We can confirm this by checking the Normal Q-Q plot, as the Shapiro-Wilks Test get´s quite sensitive in case the sample size is > 50. 

```{r}
#install.packages("ggpubr")
ggpubr::ggqqplot(penguins, "bill_length_mm", facet.by = "species")
```

The values do not deviate a lot from the reference line, thus we can assume normality and continue by checking the second assumption of the ANOVA, the variance homogeneity. 

```{r}
car::leveneTest(penguins$bill_length_mm, penguins$species)
```

The output is not significant, meaning that the variance is homogenous across the species.
As all assumptions are fulfilled we can now calculate the actual ANOVA: 

```{r}
anova_penguins <- aov(penguins$bill_length_mm ~ penguins$species)
summary(anova_penguins)

```

From the output we can conclude that there is a significant difference in bill length between the three species, but we actually don´t now which of the groups differ. Therefor we perform a post-hoc Test such as Tukey´s HSD (Honest Significant Differences):

```{r}
TukeyHSD(anova_penguins)

```



- Relations between features
- Which places or groups have similar characteristics (multivariate analysis)
- How do features change across years (time series analysis)




