---
title: "lab-exercise"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Outline

1. Get Data and Libraries
2. Introducing the dataset
3. Framing a research  question
4. Preliminary data exploration

<br>

## Get Data and Libraries

We will use data from the [`palmerpenguins`](https://github.com/allisonhorst/palmerpenguins) package for the first few sections.

```{r get data}
# install.packages("remotes")
remotes::install_github("allisonhorst/palmerpenguins")
penguins <- palmerpenguins::penguins
```


<br>

## Introducing the dataset
The palmer penguins data set contains data on the features of three different penguin species. You can have a look at the different variables using the function `head()` 

```{r}
head(penguins)
```


## Framing a research question

We want to use the palmer penguins data to show how to answer some common questions, applied in a SES context. 
The basic research question driving the exploration is: What are differences / similarities between the states (penguin species)? Are there spatio-temporal patterns / changes? 
Before starting the analysis it is always useful to do some basic data exploration / visualization exercises to get a "feeling" for your data -> How does it look like? For this we install the tidyverse packages, which includes various other packages useful for visualization (ggplot2), data wrangling (dplyr, tidyr)...

```{r}
#install.packages(tidyverse)
library(tidyverse)


```


## Preliminary data exploration

Following the steps outlined in Chapter 18 (`SES Problems and Questions`):

- Identifying key features (mean, variance, normality/distribution of the data)
- Comparison between different categories
- Relations between features
- Which places or groups have similar characteristics (multivariate analysis)
- How do features change across years (time series analysis)




